Probing internal structure of distant quasars using Microlensing

### 2014-04-06 paper1/code.tex

For the numerical computations of lightcurves for the models a and b, the microlensing code by Joachim Wambsgans was used, which is described in Wambsganss 1999. This code is based on the hierachical tree technique to calculate the gravitational effect of the lensing masses. The underlying idea is that a microlensing situation can be separated into three planes, the observer, the lens plane and the source plane (assuming the thin lens approximation to be valid). The actual effect, namely the distortion of lightrays emitted from a distant source, on the way to the observer, is numerically computed by treating this situation from the opposite side: lightrays are shot backward from the observer to lens plane, where they become deflected according to the angle, which is given by the 2-dim. mass distribution in the lens plane.

@@

For the computation of the individual deflection angles the tree idea is used. This means that the positions of all lensing masses are sorted into a 2-dim. grid (in the lens plane). This grid is subdivided into 4 smaller squares until every cell contains only one mass. For the actual calculations of the deflection angle not every mass' gravitational influence is treated equally. Instead it is made use of the fact that the influence of lenses on the considered lightray falls off with $r^{-1}$ from the point where the ray hits the lens plane. Hence masses situated at higher distances, can be grouped together in the calculation of their gravitational potential, by multipole moments. This way the amount of computation time is reduced. \\

@@

Once the map is created, the lightcurve can be obtained by specifying the transition path of the source within the map. The code thereby convolves the function describing the shape of the source with the magnification pattern of the map for each time step (see equation (6)). Note that in the most observational cases in microlensing, both lens and source are moving and further that the lens configuration, and with it the magnification pattern, is changing with time. While the first subtlety is taken care of by a coordinate transformation in this analysis, for the second one the lens configuration is assumed to be constant in time.  \\\\
For the model a it was desirable to work with a caustic geometry as simple as possible. Therefore only two equal point like masses were used as lenses, to produce the magnification map depicted in figure (...). For a description of the parameter set which characterizes the map see (reference). The magnification maps calculated by the microlensing code are in units of Einstein radii, hence they are principally scale free. Physical scale is introduced to them by multiplying the pixels by the correct distance factor. The Einstein angle is given by   

@@

For the next step, the computation of the actual lightcurve, the code was modified to also allow for crescent shaped images specified through the parameter set $R_p,R_n,a,b$. Here $R_p$ denotes the outer radius of the crescent and $R_n$ the inner one. The orientation of the source image with respect to the magnification map is specified by the parameters $a$ and $b$ as the shift of the center of the inner disk from the center of the outer disk in $x$- and $y$-direction respectively. In the original version of the code, gaussian and disk shaped images where already implemented. Those are completely characterized by the single parameter $R_p$. The values of the parameters are specified in pixel units corresponding to the magnification map. Further one needs to specify the start and end point coordinates of the path, which the center of the source image follows through the magnification map (see the depiction in figure (...)). Hence the points along this path are specified through the number of timesteps for which the computation of the brightness is to be carried out. Those points correspond to the actual measurement of the brightness of an object in the observational case. For each timestep the two-dimensional convolution of equation (6) is carried out numerically for the position of the source on the magnification map. \\

>>>>

For the numerical computations of lightcurves for the models a and b, the microlensing code by Joachim Wambsgans was used, which is described in \cite{1999A&A...346L...5W}. This code is based on the hierachical tree technique to calculate the gravitational effect of the lensing masses. The underlying idea is that a microlensing situation can be separated into three planes, the observer, the lens plane and the source plane (assuming the thin lens approximation to be valid). The actual effect, namely the distortion of lightrays emitted from a distant source, on the way to the observer, is numerically computed by treating this situation from the opposite side: lightrays are shot backward from the observer to lens plane, where they become deflected according to the angle, which is given by the 2-dim. mass distribution in the lens plane.

@@

For the computation of the individual deflection angles the tree idea is used. This means that the positions of all lensing masses are sorted into a 2-dimensional grid (in the lens plane). This grid is subdivided into 4 smaller squares until every cell contains only one mass. For the actual calculations of the deflection angle not every mass gravitational influence is treated equally. Instead it is made use of the fact that the influence of lenses on the considered lightray falls off with $r^{-1}$ from the point where the ray hits the lens plane. Hence masses situated at higher distances, can be grouped together in the calculation of their gravitational potential, by multipole moments. This way the amount of computation time is reduced. \\

@@

Once the map is created, the lightcurve can be obtained by specifying the transition path of the source within the map. The code thereby convolves the function describing the shape of the source with the magnification pattern of the map for each time step (see equation \ref{eqn:ft2d}). Note that in the most observational cases in microlensing, both lens and source are moving and further that the lens configuration, and with it the magnification pattern, is changing with time. While the first subtlety is taken care of by a coordinate transformation in this analysis, for the second one the lens configuration is assumed to be constant in time.  \\\\
For the model a it was desirable to work with a caustic geometry as simple as possible. Therefore only two equal point like masses were used as lenses, to produce the magnification map depicted in figure \ref{fig:magnification}. For a description of the parameter set which characterizes the map see (reference). The magnification maps calculated by the microlensing code are in units of Einstein radii, hence they are principally scale free. Physical scale is introduced to them by multiplying the pixels by the correct distance factor. The Einstein angle is given by   

@@

For the next step, the computation of the actual lightcurve, the code was modified to also allow for crescent shaped images specified through the parameter set $R_p,R_n,a,b$. Here $R_p$ denotes the outer radius of the crescent and $R_n$ the inner one. The orientation of the source image with respect to the magnification map is specified by the parameters $a$ and $b$ as the shift of the center of the inner disk from the center of the outer disk in $x$- and $y$-direction respectively. In the original version of the code, gaussian and disk shaped images where already implemented. Those are completely characterized by the single parameter $R_p$. The values of the parameters are specified in pixel units corresponding to the magnification map. Further one needs to specify the start and end point coordinates of the path, which the center of the source image follows through the magnification map (see the depiction in figure \ref{fig:magnification}). Hence the points along this path are specified through the number of timesteps for which the computation of the brightness is to be carried out. Those points correspond to the actual measurement of the brightness of an object in the observational case. For each timestep the two-dimensional convolution of equation \ref{eqn:ft2d} is carried out numerically for the position of the source on the magnification map. \\

### 2014-04-06 paper1/intro.tex

\section{introduction}
More than half a century after their discovery, quasars still maintain the interest of a productive part of the astrophysical community. Although several mechanisms have been proposed 
to explain how such systems function a single mechanism is widely accepted at the present time. The model was first proposed by Salpeter in 1964 (citation), folowed by Zel'dovich and Novikov in the same year (citation) and Lynden-Bell five years later. The present name of the model is the Super-Massive Black Hole (SMBH) paradigm.\\

@@

Through extensive observations done in the past 50 years, an important part of the assumptions and details  regarding the mechanism have been coomfirmed and uncovered. Still, the black hole shadow and the surrounding luminous acrretting material (the black hole silouette) remains to be probed. Advancements in the very long baseline interferometer observations have made possible since 2007 the detection of structures of a few Schwarzschild radii at the SgrA* object at the center of our galaxy (Doeleman 2008). Furthermore, jet launching structure near the supermasve black hole in M87 were resolved with the same observation tool (Doeleman 2012 Jet launching). With all the advancement, the current number of baselines does not allow the direct creation of images from the observations still it does allow the fitting of the date to predefined models(Kamruddin 2013).
A whole range of models have been applied starting from simple geometric models to more complex, physically sensible models (Doeleman et al 2008, Fish et al 2011, Broderick et al 2011, Moscibrodzka et al 2009, Dexter 2010 and others). A significant fraction of the later models predict a crescent shaped silouette of the black hole. This motivated Kamruddin and Dexter (2013) to use a simple geometric crescent model to fit the data. The succesfull results encouraged the authors to speculate that the shadow of the black hole will be observed in the near fututre with the previously mentioned observation method. \\

Hundreds of thousands of quasars have been detected up to date, mainly from the Sloan Digital Sky Survey. The great majority of them present a redshift larger than 2.15 of their spectra (Paris et al 20Nov 2013 SDSS catalogue). Therefore the typical distance to the objects is in the order of gigaparsecs or above. A value that is orders of magnitude 
larger than the distance to M87, estimated to be 50 Mpc? (citation). Direct observation of the black hole silouette of quasars require significant technological advances. We argue in the present paper
 that one can probe and study the black hole shadow and its proximal environment belonging to a quasar without having to resolve sub-event horizon-scales. We consider that this is possible by analyzing the data from microlensing-events.  Moreover, unlike with the few nearby AGNs, the study of the very large quasar population allows for the statistical refinement of the global properties of AGN by highlighting any particular
characteristic that is present only in SGA* and M87 central region. Furthermore, by probing the event-horizon-scale inner regions of of a large number of objects will facilitate the study of the evolution of the previously menttioned region.

Gravitational microlensing of QSOs have been predicted as early as 1981 when Gott suggested (Gott 1981) that if haloes of galaxies were composed of stars with masses less than `` 0.1 $M_{sol}$, then these stars acting as individual lenses whould produce fluctuations of the order of 
unity in the intensities of the QSO images on time scales of 1 -14 years '' (Gott 1981). What creates the magnification event is the relative and independent motion of the lens and source with respect to the observer. Since the respective lens is surrounded by similar objects one would expect to observe during a long enough time interval
multiple microlensing events of the same source, the QSO. The sequential alignment of the source and observer with different lensing objects will result in an aparently random increase of the QSO light flux analogous to the scintilation events in a meter of radioactivity.

Estimates of the Einstein radius for the particular case of a QSO with redshift 2 lensed by a compact stellar mass object with redshift 0.5 was found to be $10^-6 \sqrt{M/M_{sol}} arcsec$ (citation Wambsganss Quasar microlensing 527). The Schwarzschil radius of the Sgr A* black hole, situated at 8 kpc from Earth [Reid M.J. 1993 The distance to the center of the galaxy] spans an angle of 10 microarcsec on the sky [Doeleman 2008]. If the quasars observed at z =2 would have similar linear sizes, the aparent size of the  luminous part of the QSO would be many orders of magnitude below the 1 microarcsec level. Therefore the aparent size would be significantly smaller than the corresponding Einstein radius of a solar mass lens. 
The magnitude differences between the Einstein angle and the aparrent size of the bright part of the QSO motivates the introduction in the first part of the paper of simplified models for     
calculating the lightcurves of QSO magnification events. The simplified models are introduced in the theory section of the present paper. In the theory (II) section of the present article with an introduction to the basic microlensing theory in which we present the general microlensing equation followed by the concept of magnification. We underline that for the particular case of micro gravitational lensing
the images are not resolved and thus the magnifications of each images contribute to a total magnification. Furthermore, the concept of critical curves and caustics are introduced. Afterwards we focus on a particular singularity, namely the fold singularity. We choose to study events related to the previously mentioned singularity since the magnification events caused by the crossing of such boundaries are most probable. A quick inspection of figure 1 by the reader should motivatethe previous statement. 
In continuation we introduce the commonly used result involving the magnification of a point-like source near a fold singularity. Namely that the magnification is aproximately linear outside the caustic and decreases from infinity with the square root of the distance to the boundary when inside the caustic.  
In the last section of the Introduction (II) we present the general equations for estimating the flux received from an extended source near a fold. The only elements required to compute the flux as a function of time are the two dimensional brightness function of the object and the magnification of a point-like source. One of the core
assumptions made when developing the simple models is that the einstein angle is orders of magnitude greater than the apparent size of the source. Due to this assumption one can imagine the caustic as an infinite wall to be crossed by the source as presented in figure 2. The consequence of the limit is that the information regarding the profiles of the source along any direction parallel to the boundary value is reduced to the corresponding integral value. The effect allows the introduction of a one dimensional flux function, which is the two dimensional flux integrated along the axis paralel to the boundary. The one dimensional profile contains exhaustively all the information that can be extracted  from the lens generated light-curve \\

In the third section we introduce the three source profiles used in our models and simulations. We start with two commonly used and simple sources. Namely the gaussian brightness distribution and constant intensity disk. In addition we use a geometric crescent-shaped source effectively the same as the one introduced by (geometric crescent paper) with constant surface brightness. \\
The fourth section contains the results of our simplified models calculations, including the numerical equation for lightcurve calculations and the corresponding plots. In order to facilitate comparison bewteen the lightcurves of the three distinct sources, we constrain the luminosity and half-light radius of all sources to have the same value.

>>>>

\section{introduction}
More than half a century after their discovery, quasars still maintain the interest of a productive part of the astrophysical community. Although several mechanisms have been proposed 
to explain how such systems function, a single mechanism is widely accepted at the present time. The model was first proposed by \cite{1964ApJ...140..796S}, followed by \cite{1964SPhD....9..246Z} in the same year and \cite{1969Natur.223..690L} five years later. The present name of the model is the Super-Massive Black Hole (SMBH) paradigm.\\

@@

Through extensive observations done in the past 50 years, an important part of the assumptions and details  regarding the mechanism have been confirmed and uncovered. Still, the black hole shadow and the surrounding luminous accreting material (the black hole silouette) remains to be probed. Advancements in the very long baseline interferometer observations have made possible since 2007, the detection of structures of a few Schwarzschild radii at the Sagitarius A object at the center of our galaxy \citep{2008JPhCS.131a2055D}. Furthermore, jet launching structure near the supermasve black hole in M87 were resolved with the same observation tool \citep{2012Sci...338..355D}. With all the advancement, the current number of baselines does not allow the direct creation of images from the observations still it does allow the fitting of the data to predefined models \citep{2013MNRAS.434..765K}.

A whole range of models have been applied starting from simple geometric models to more complex, physically sensible models (Doeleman et al 2008, Fish et al 2011, Broderick et al 2011, Moscibrodzka et al 2009, Dexter 2010 and others). A significant fraction of the later models predict a crescent shaped silouette of the black hole. This motivated \cite{2013MNRAS.434..765K} to use a simple geometric crescent model to fit the data. The succesfull results encouraged the authors to speculate that the shadow of the black hole will be observed in the near fututre with the previously mentioned observation method. \\

Hundreds of thousands of quasars have been detected up to date, mainly from the Sloan Digital Sky Survey. The great majority of them present a redshift larger than 2.15 of their spectra (Paris et al 20Nov 2013 SDSS catalogue). Therefore the typical distance to the objects is in the order of Gigaparsecs (Gpc) or above. A value that is orders of magnitude 
larger than the distance to M87, estimated to be 50 Mpc? (citation). Direct observation of the black hole silouette of quasars require significant technological advances. We argue in the present paper
 that one can probe and study the black hole shadow and its proximal environment belonging to a quasar without having to resolve sub-event horizon scales. We consider that this is possible by analyzing the data from microlensing-events.  Moreover, unlike with the few nearby AGNs, the study of the very large quasar population allows for the statistical refinement of the global properties of AGN by highlighting any particular
characteristic that is present only in SGA* and M87 central region. Furthermore, by probing the event-horizon-scale inner regions of of a large number of objects will facilitate the study of the evolution of the previously mentioned region.

Gravitational microlensing of QSOs have been predicted as early as 1981 when \cite{1981ApJ...243..140G} suggested that ''if haloes of galaxies were composed of stars with masses less than  0.1 $M_{sol}$, then these stars acting as individual lenses whould produce fluctuations of the order of 
unity in the intensities of the QSO images on time scales of 1 -14 years '' (Gott 1981). What creates the magnification event is the relative and independent motion of the lens and source with respect to the observer. Since the respective lens is surrounded by similar objects one would expect to observe during a long enough time interval
multiple microlensing events of the same source, the QSO. The sequential alignment of the source and observer with different lensing objects will result in an aparently random increase of the QSO light flux analogous to the scintilation events in a meter of radioactivity.

Estimates of the Einstein radius for the particular case of a QSO at redshift 2 lensed by a compact stellar mass object at redshift 0.5 was found to be $10^-6 \sqrt{M/M_{sol}} arcsec$ \citep{2001PASA...18..207W}. The Schwarzschild radius of the Sgr A* black hole, situated at 8 kpc from Earth \citep{1993ARA&A..31..345R} spans an angle of 10 microarcsec on the sky \citep{2008JPhCS.131a2055D}. If the quasars observed at z =2 would have similar linear sizes, the aparent size of the  luminous part of the QSO would be many orders of magnitude below the 1 microarcsec level. Therefore the aparent size would be significantly smaller than the corresponding Einstein radius of a solar mass lens. 
The magnitude differences between the Einstein angle and the aparrent size of the bright part of the QSO motivates the introduction in the first part of the paper of simplified models for     
calculating the lightcurves of QSO magnification events. The simplified models are introduced in the theory section of the present paper. In the theory (II) section of the present article with an introduction to the basic microlensing theory in which we present the general microlensing equation followed by the concept of magnification. We underline that for the particular case of micro gravitational lensing,
the images are not resolved and thus the magnifications of each images contribute to a total magnification. Furthermore, the concept of critical curves and caustics are introduced. Afterwards we focus on a particular singularity, namely the fold singularity. We choose to study events related to the previously mentioned singularity since the magnification events caused by the crossing of such boundaries are most probable. A quick inspection of figure 1 by the reader should motivate the previous statement. 
In continuation we introduce the commonly used result involving the magnification of a point-like source near a fold singularity. Namely that the magnification is aproximately linear outside the caustic and decreases from infinity with the square root of the distance to the boundary when inside the caustic.  
In the last section of the Introduction (II) we present the general equations for estimating the flux received from an extended source near a fold. The only elements required to compute the flux as a function of time are the two dimensional brightness function of the object and the magnification of a point-like source. One of the core
assumptions made when developing the simple models is that the einstein angle is orders of magnitude greater than the apparent size of the source. Due to this assumption one can imagine the caustic as an infinite wall to be crossed by the source as presented in figure 2. The consequence of the limit is that the information regarding the profiles of the source along any direction parallel to the boundary value is reduced to the corresponding integral value. The effect allows the introduction of a one dimensional flux function, which is the two dimensional flux integrated along the axis paralel to the boundary. The one dimensional profile contains exhaustively all the information that can be extracted  from the lens generated light-curve \\

In the third section we introduce the three source profiles used in our models and simulations. We start with two commonly used and simple sources. Namely the gaussian brightness distribution and constant intensity disk. In addition we use a geometric crescent-shaped source effectively the same as the one introduced by \citep{2013MNRAS.434..765K} with constant surface brightness. \\
The fourth section contains the results of our simplified models calculations, including the numerical equation for lightcurve calculations and the corresponding plots. In order to facilitate comparison bewteen the lightcurves of the three distinct sources, we constrain the luminosity and half-light radius of all sources to have the same value.
